---
title: "Match collectors to Wikidata"
author: "Mathias Dillen"
date: "05/13/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,eval = FALSE)
```

## Packages

```{r libs}
library(tidyverse)
library(jsonlite)
library(data.table)
library(magrittr)
```

## Functions

Query Wikidata, based on the WikidataQueryServiceR package functions. For use, please change the user agent to a project name or your Wikidata account username.

```{r queryWiki}
querki <- function(query,h="text/csv",agent="insert_your_wiki_name_here_please") {
  require(httr)
  response <- httr::GET(url = "https://query.wikidata.org/sparql", 
                        query = list(query = query),
                        httr::add_headers(Accept = h),
                        httr::user_agent(agent))
  return(httr::content(response,
                       type=h,
                       col_types = cols(.default = "c")))
}
```

Match (a combination of) names to Wikidata item labels. This function requires  a list of wikidata item `id`s and `itemLabel`s. It tries to match names from a list of person records with first name `FNAME`, middle name(s) `MNAME` and last name `LNAME`. First and middle names are optional, but need to be present in the dataset (as `NA` if unavailable). The function returns the matches as concatenated strings of `id`s and of `itemLabel`s separated by `|`.

The function is used as part of the matching script. It also expects in the Wikidata item list variables for `surname`, the last word of the Wikidata item label, and for `initials`, the initials of the item label. These variables are readily created from the query results below.

```{r matchName}
matchName <- function(wikidataItems,specimenCollector) {
  #Requirements:
  ##specimenCollector$LNAME (last name)
  ##wikidataItems$surname (extracted from Wikidata item label)
  ##wikidataItems$itemLabel (Wikidata item label)
  ##wikidataItems$id (Wikidata item ID)
  ##WikidataItems$initials (initials of Wikidata item label)
  #Requirements, but can be NA:
  ##specimenCollector$FNAME (first name)
  ##specimenCollector$MNAME (middle names)
  
  #Try to fuzzy match a known last name into the last name of the Wiki label
  wikidataItemsFiltered = wikidataItems %>%  
    filter(agrepl(specimenCollector$LNAME,
                  surname))
  #Ensure they're of similar length (max 1 char longer)
  wikidataItemsFiltered %<>% 
    filter(nchar(specimenCollector$LNAME) > nchar(surname)-2)
  len = dim(wikidataItemsFiltered)[1]
  
  #Do extra fuzzy matching for first name
  ##Fuzzy matching if no abbreviations used
  ##Exact matching if abbreviated (with a period)
  if (!is.na(specimenCollector$FNAME)) {
    if (!grepl(".",specimenCollector$FNAME,
               fixed=T)) {
      wikidataItemsFiltered %<>% 
        filter(agrepl(specimenCollector$FNAME,
                      itemLabel))
      len = dim(wikidataItemsFiltered)[1]
    }
    if (grepl(".",specimenCollector$FNAME,
              fixed=T)) {
      wikidataItemsFiltered %<>% 
        filter(grepl(
          gsub("\\.",
               "",
               specimenCollector$FNAME),
          initials))
      len=dim(wikidataItemsFiltered)[1]
    }
  }
  
  #Fuzzy matching with middle name only if multiple matches were found
  if (len>1&
      !is.na(specimenCollector$MNAME)) {
    wikidataItemsFiltered %<>% 
      filter(agrepl(specimenCollector$MNAME,
                    itemLabel))
    len = dim(wikidataItemsFiltered)[1]
  }
  return(c(len,
           paste(wikidataItemsFiltered$itemLabel,collapse="|"),
           paste(wikidataItemsFiltered$id,collapse="|")))
}
```

## Query wikidata

Retrieve results of SPARQL queries in a CSV format. Each query returns Wikidata item ids and certain other properties, such as different properties for external identifiers and dates of birth/death, but also including floruit dates, the English item label and alternative labels. All items in these queries need to be instances of `P31` human `Q5`.

The queries may take a few seconds and may occasionally time out.

# Initialize

```{r startQuery}
raw = list()
iter = 1
```


# Entomologists of the World ID

```{r entom}
query <- 'SELECT DISTINCT ?item ?itemLabel ?itemAltLabel ?entom_id ?orcid ?viaf ?isni ?yob ?yod ?fly ?wyb ?wye WHERE {
  ?item wdt:P31 wd:Q5 .
  ?item wdt:P5370 ?entom_id. #entomologists of the world .
  SERVICE wikibase:label { bd:serviceParam wikibase:language "[AUTO_LANGUAGE],en" } 
  OPTIONAL { ?item wdt:P496 ?orcid .}
  OPTIONAL { ?item wdt:P214 ?viaf .}
  OPTIONAL { ?item wdt:P213 ?isni .}
  OPTIONAL { ?item wdt:P569 ?dob . BIND(YEAR(?dob) as ?yob) }
	OPTIONAL { ?item wdt:P570 ?dod . BIND(YEAR(?dod) as ?yod) }
	OPTIONAL { ?item wdt:P1317 ?fl . BIND(YEAR(?fl) as ?fly) }
	OPTIONAL { ?item wdt:P2031 ?wpb . BIND(YEAR(?wpb) as ?wyb) }
	OPTIONAL { ?item wdt:P2032 ?wpe . BIND(YEAR(?wpe) as ?wye) }
}'
raw[[iter]] = querki(query)
iter = iter + 1
```

# BHL Creator ID

```{r bhl}
query <- 'SELECT DISTINCT ?item ?itemLabel ?itemAltLabel ?bhl_id ?orcid ?viaf ?isni ?yob ?yod ?fly ?wyb ?wye WHERE {
  ?item wdt:P31 wd:Q5 .
  ?item wdt:P4081 ?bhl_id. #BHL creator .
  SERVICE wikibase:label { bd:serviceParam wikibase:language "[AUTO_LANGUAGE],en" } 
  OPTIONAL { ?item wdt:P496 ?orcid .}
  OPTIONAL { ?item wdt:P214 ?viaf .}
  OPTIONAL { ?item wdt:P213 ?isni .}
  OPTIONAL { ?item wdt:P569 ?dob . BIND(YEAR(?dob) as ?yob) }
	OPTIONAL { ?item wdt:P570 ?dod . BIND(YEAR(?dod) as ?yod) }
	OPTIONAL { ?item wdt:P1317 ?fl . BIND(YEAR(?fl) as ?fly) }
	OPTIONAL { ?item wdt:P2031 ?wpb . BIND(YEAR(?wpb) as ?wyb) }
	OPTIONAL { ?item wdt:P2032 ?wpe . BIND(YEAR(?wpe) as ?wye) }
}'
raw[[iter]] = querki(query)
iter = iter + 1
```

# Harvard index of botanists ID

```{r harv}
query <- 'SELECT DISTINCT ?item ?itemLabel ?itemAltLabel ?harv_id ?orcid ?viaf ?isni ?yob ?yod ?fly ?wyb ?wye WHERE {
  ?item wdt:P31 wd:Q5 .
  ?item wdt:P6264 ?harv_id. #Harvard index of botanists .
  SERVICE wikibase:label { bd:serviceParam wikibase:language "[AUTO_LANGUAGE],en" } 
  OPTIONAL { ?item wdt:P496 ?orcid .}
  OPTIONAL { ?item wdt:P214 ?viaf .}
  OPTIONAL { ?item wdt:P213 ?isni .}
  OPTIONAL { ?item wdt:P569 ?dob . BIND(YEAR(?dob) as ?yob) }
	OPTIONAL { ?item wdt:P570 ?dod . BIND(YEAR(?dod) as ?yod) }
	OPTIONAL { ?item wdt:P1317 ?fl . BIND(YEAR(?fl) as ?fly) }
	OPTIONAL { ?item wdt:P2031 ?wpb . BIND(YEAR(?wpb) as ?wyb) }
	OPTIONAL { ?item wdt:P2032 ?wpe . BIND(YEAR(?wpe) as ?wye) }
}'
raw[[iter]] = querki(query)
iter = iter + 1
```

# Zoobank ID

```{r zoo}
query <- 'SELECT DISTINCT ?item ?itemLabel ?itemAltLabel ?zoo_id ?orcid ?viaf ?isni ?yob ?yod ?fly ?wyb ?wye WHERE {
  ?item wdt:P31 wd:Q5 .
  ?item wdt:P2006 ?zoo_id. #zoobank .
  SERVICE wikibase:label { bd:serviceParam wikibase:language "[AUTO_LANGUAGE],en" } 
  OPTIONAL { ?item wdt:P496 ?orcid .}
  OPTIONAL { ?item wdt:P214 ?viaf .}
  OPTIONAL { ?item wdt:P213 ?isni .}
  OPTIONAL { ?item wdt:P569 ?dob . BIND(YEAR(?dob) as ?yob) }
	OPTIONAL { ?item wdt:P570 ?dod . BIND(YEAR(?dod) as ?yod) }
	OPTIONAL { ?item wdt:P1317 ?fl . BIND(YEAR(?fl) as ?fly) }
	OPTIONAL { ?item wdt:P2031 ?wpb . BIND(YEAR(?wpb) as ?wyb) }
	OPTIONAL { ?item wdt:P2032 ?wpe . BIND(YEAR(?wpe) as ?wye) }
}'
raw[[iter]] = querki(query)
iter = iter + 1
```

#IPNI ID

```{r ipni}
query <- 'SELECT DISTINCT ?item ?itemLabel ?itemAltLabel ?ipni_id ?orcid ?viaf ?isni ?yob ?yod ?fly ?wyb ?wye WHERE {
  ?item wdt:P31 wd:Q5 .
  ?item wdt:P586 ?ipni_id. #ipni
  SERVICE wikibase:label { bd:serviceParam wikibase:language "[AUTO_LANGUAGE],en" } 
  OPTIONAL { ?item wdt:P496 ?orcid .}
  OPTIONAL { ?item wdt:P214 ?viaf .}
  OPTIONAL { ?item wdt:P213 ?isni .}
  OPTIONAL { ?item wdt:P569 ?dob . BIND(YEAR(?dob) as ?yob) }
	OPTIONAL { ?item wdt:P570 ?dod . BIND(YEAR(?dod) as ?yod) }
	OPTIONAL { ?item wdt:P1317 ?fl . BIND(YEAR(?fl) as ?fly) }
	OPTIONAL { ?item wdt:P2031 ?wpb . BIND(YEAR(?wpb) as ?wyb) }
	OPTIONAL { ?item wdt:P2032 ?wpe . BIND(YEAR(?wpe) as ?wye) }
}'
raw[[iter]] = querki(query)
iter = iter + 1
```

# Wikispecies article

```{r wsp}
query <- 'SELECT DISTINCT ?item ?itemLabel ?itemAltLabel ?article ?orcid ?viaf ?isni ?yob ?yod ?fly ?wyb ?wye WHERE {
	?item wdt:P31 wd:Q5 .
  ?article 	schema:about ?item ;
			schema:isPartOf <https://species.wikimedia.org/> .
	SERVICE wikibase:label { bd:serviceParam wikibase:language "[AUTO_LANGUAGE],en" }
	OPTIONAL { ?item wdt:P496 ?orcid .}
  OPTIONAL { ?item wdt:P214 ?viaf .}
  OPTIONAL { ?item wdt:P213 ?isni .}
  OPTIONAL { ?item wdt:P569 ?dob . BIND(YEAR(?dob) as ?yob) }
	OPTIONAL { ?item wdt:P570 ?dod . BIND(YEAR(?dod) as ?yod) }
	OPTIONAL { ?item wdt:P1317 ?fl . BIND(YEAR(?fl) as ?fly) }
	OPTIONAL { ?item wdt:P2031 ?wpb . BIND(YEAR(?wpb) as ?wyb) }
	OPTIONAL { ?item wdt:P2032 ?wpe . BIND(YEAR(?wpe) as ?wye) }
}'
raw[[iter]] = querki(query)
iter = iter + 1
```

# Join all results

Join all results and filter the result for unique Wikidata ids. Some may be duplicated if they have multiple identifiers for the same property (e.g. multiple IPNI identifiers for a single person).

```{r join}
#full join
wikiResults = full_join(raw[[1]],
                        raw[[2]])
if (length(raw)>2) {
  for (i in 3:length(raw)) {
    wikiResults = full_join(wikiResults,
                            raw[[i]])
  }
}

#unique items, add id
wikiResults.items = filter(wikiResults,!duplicated(item))
wikiResults.items$id = gsub("http://www.wikidata.org/entity/",
                            "",
                            wikiResults.items$item)
```

# Add the (presumed) surname and initials

Additional variables for the (presumed) surname and the initials of the name are calculated.

```{r wikiResultsX}
#add surname (last word of label)
wikiResults.items$surname = gsub("^(.*[\\s])",
                                 "",
                                 wikiResults.items$itemLabel,
                                 perl=T)

#create initials for the wikidata item labels
#also remove quotes
wikiResults.items$initials = gsub("\'",
                              "",
                              wikiResults.items$itemLabel,
                              fixed=T)
wikiResults.items$initials = gsub("\"",
                              "",
                              wikiResults.items$initials,
                              fixed=T)
wikiResults.items$initials = gsub("(?<!\\s).",
                              "",
                              wikiResults.items$initials,
                              perl=T)
wikiResults.items$initials = paste0(
  substr(wikiResults.items$itemLabel,1,1),
  wikiResults.items$initials)
```

#Aliases

The aliases are listed along the default labels in a separate table. They will only be used for exact matching.

```{r altNames}
#unlist the concatenated aliases (with , as a separator)
#uses data.table
wikiResults.altnames = 
  setDT(wikiResults.items)[, list(itemAltLabel = 
                                       unlist(str_split(itemAltLabel,
                                                        pattern=", "))),
                              .(id,itemLabel)]

#take the alt labels and set them as itemLabel
wikiResults.altnames2 = wikiResults.altnames %>% 
  select(-itemLabel) %>%
  rename(itemLabel = itemAltLabel)

#then stack the default labels and aliases into one table
wikiResults.altnames = rbind(wikiResults.altnames2,
                             select(wikiResults.altnames,
                                    -itemAltLabel))

#Remove duplicate id-label pairs
wikiResults.altnames$test = paste0(wikiResults.altnames$id,
                              wikiResults.altnames$itemLabel)
wikiResults.altnames %<>% 
  filter(!duplicated(test)) %>%
  select(-test)

#remove missing labels
wikiResults.altnames %<>%
  filter(!is.na(itemLabel))
```


## Import list of collectors to match

Example in this case used for the COST Mobilise STSM is an export from the collector table of Meise Botanic Garden's collection management system, BG-Base.

```{r bgbimp}
bgb = read_tsv("data/COLLECTORS_28OCT19.TXT")
```

Extract year from the dates of birth/death and concatenate the full name based on FNAME, MNAME and LNAME (i.e. first, middle and last names).

```{r bgbprep}
#extract year
for (i in 1:dim(bgb)[1]) {
  bgb$birthDate[i] = substr(bgb$BIRTH_DT[i],
                        nchar(bgb$BIRTH_DT[i])-3,
                        nchar(bgb$BIRTH_DT[i]))
  bgb$deathDate[i] = substr(bgb$DEATH_DT[i],
                        nchar(bgb$DEATH_DT[i])-3,
                        nchar(bgb$DEATH_DT[i]))
}

#compose fullname
bgb$fullname = paste(bgb$FNAME,
                     bgb$MNAME,
                     bgb$LNAME)
bgb$fullname = gsub("NA ",
                    "",
                    bgb$fullname)
```

Attach `minSpecDate` and `maxSpecDate` to indicate a range of dates when the person was alive as specimens were supposedly collected by them at that time (floruit date). If the range between these dates is greater than 100, it's considered unreliable and the dates are removed. Also attach `n`, the number of specimens linked to them.

```{r dateX}
exp = read_tsv("data/MeiseBG-specimens.tsv")
bgb$n = NA
bgb$minSpecDate = NA
bgb$maxSpecDate = NA
speccoll = count(exp,COLL_ID)
for (i in 1:dim(bgb)[1]) {
  k = speccoll[speccoll$COLL_ID==bgb$COLL_ID[i],]
  if (is.na(bgb$birthDate[i])) {
    dates = filter(exp,
                   bgb$COLL_ID[i]==COLL_ID,
                   is.na(COLL_DT)==F
                   )$COLL_DT
    if (length(dates)>0) {
      bgb$minSpecDate[i] = min(as.numeric(substr(dates,8,11)))
      bgb$maxSpecDate[i] = max(as.numeric(substr(dates,8,11)))
    }
  }
  if (is.na(bgb$minSpecDate[i])==F&
      is.na(bgb$maxSpecDate[i])==F) {
    dur = bgb$maxSpecDate[i] - bgb$minSpecDate[i]
    if (dur > 100) {
      bgb$minSpecDate[i] = NA
      bgb$maxSpecDate[i] = NA
    }
  }
  bgb$n[i]=k$n[1]
  if (i%%100==0) {
    print(i)
  }
}
```

Only try to match names of people who collected a specimen and for which we have some sort of date. Also, remove the records without any name. For the floruit date, we assume a childhood period of 15 years, so we adjust the minimum date with that.

```{r filterBGB}
#only people who have collected specimens
bgb2 = filter(bgb,n>0)

#only those who have some sort of date
bgb3 = filter(bgb2,!is.na(birthDate)|
                !is.na(minSpecDate))
bgb3 = filter(bgb3,!is.na(LNAME)) #remove no collector specified

#as numeric for calculations
bgb3$birthDate = as.numeric(bgb3$birthDate)
bgb3$deathDate = as.numeric(bgb3$deathDate)

#add 15y margin for childhood
bgb3$minSpecDate2 = bgb3$minSpecDate - 15
```

## Additional date info for Wikidata items

A floruit range of `floruitDate1` and `floruitDate2` is compiled based on the floruit dates or date ranges taken from the SPARQL query results.

```{r addDate}
#add date ranges to wikidata items based on floruit dates
#if no year of birth is available
wikiResults.noyob = filter(wikiResults.items,is.na(yob))
wikiResults.noyob$floruitDate1 = NA
wikiResults.noyob$floruitDate2 = NA

#Wikidata items with no year of birth, but with some floruit year
wikiResults.floruit = filter(wikiResults.noyob,
                 !is.na(fly)|
                   !is.na(wyb)|
                   !is.na(wye))

for (i in 1:dim(wikiResults.floruit)[1]) {
  if (!is.na(wikiResults.floruit$fly[i])) {
    wikiResults.floruit$floruitDate1[i] = wikiResults.floruit$fly[i]
    wikiResults.floruit$floruitDate2[i] = wikiResults.floruit$fly[i]
  }
  if (is.na(wikiResults.floruit$fly[i])) {
    wikiResults.floruit$floruitDate1[i] = wikiResults.floruit$wyb[i]
    wikiResults.floruit$floruitDate2[i] = wikiResults.floruit$wye[i]
  }
}

```

## Matching

# Inititalize the matching

The matching process will provide the following output:

- `options`: Possible label strings of Wikidata items matching the collector. Separated by `|`.
- `lopt`: Number of Wikidata matches found and withheld.
- `wdid`: Wikidata item IDs of the matched strings in `options`.
- `score`: A descriptive indicator of how the match was determined.

```{r matchInit}
bgb3$options = NA #possible labels that match, separated by pipe
bgb3$lopt = NA #number of matches
bgb3$wdid = NA #to put wikidata item ids, separated by pipes
bgb3$score = "" #some sort of indicator how the matching happened

##if using other source of data, ensure that:
#o bgb3$fullname = full concatenated string of the name, in order from first to last name
#can contain initials with dots
#o bgb3$birthDate = year of birth
#o bgb3$deathDate = year of death
#o bgb3$LNAME = last name
#o bgb3$FNAME = first name [optional]
#o bgb3$MNAME = middle name(s) [optional]
#o bgb3$minSpecDate2 = earliest year a specimen was collected by this person (-15y to account for childhood)
#o bgb3$maxSpecDate = latest year a specimen was collected by this person
#NA if not available
```

# Matching loop

This may take a couple of minutes.

1. Exact matches
This matching script first tries to identify any exact string matches between the full name and the full label. If multiple such matches are found, it adds an extra filter based on year of birth (e.g. fathers and sons with the same name). Either way, these matches are withheld.

2. Fuzzy matches
Here, the matchName function from above is applied. To streamline the fuzzy matches, a pre-filtering is done based on the dates of birth/death/floruit.

2.1. Year of birth/death
First, a matching attempt is made with those Wikidata items which have the exact same year of birth or death. If any records are found and the fuzzy matching is successful, the script skips to the next record immediately.

2.2. If no birth/death match was found or the fuzzy matching was not successful, the script tries to filter the Wikidata floruit dates instead. The floruit dates need to realistically fit the lifespan indicated in the list of specimen collectors.

2.3. If still no match is found, the inferred floruit range from the collected specimens is used to filter the Wikidata items based on their year of birth/death.

Matching floruit dates on both sides was not implemented.

```{r matchLoop}
for (i in 1:dim(bgb3)[1]) {
  #Exact matches
  truem = filter(wikiResults.altnames,
                 itemLabel==bgb3$fullname[i])
  if (dim(truem)[1]>1) {
    truem %<>%
      group_by(id) %>%
      summarise(itemLabel = paste(itemLabel,
                                  collapse=","))
    #if more than one exact match, try to filter on birth date
    if (!is.na(bgb3$birthDate[i])) {
      truem2 = filter(wikiResults.items,
                      id%in%truem$id,
                      (yob==bgb3$birthDate[i]|is.na(yob)))
      truem %<>%
        filter(id%in%truem2$id)
      if (dim(truem2)[1]>0) {
        bgb3$lopt[i] = dim(truem)[1]
        bgb3$score[i] = paste0(bgb3$score[i],
                               "Multiple Matches")
        bgb3$options[i] = paste(truem$itemLabel,
                                collapse="|")
        bgb3$wdid[i] = paste(truem$id,
                             collapse="|")
        next
      }
    }
  }
  
  #Otherwise, just list the exact matches
  if (dim(truem)[1]!=0) {
    bgb3$lopt[i] = dim(truem)[1]
    bgb3$score[i] = paste0(bgb3$score[i],
                           "Single Match")
    bgb3$options[i] = paste(truem$itemLabel,
                            collapse="|")
    bgb3$wdid[i] = paste(truem$id,
                         collapse="|")
    next
  }
  #if no exact name match, first try to find birth or death year matches
  bgb3$score[i] = paste0(bgb3$score[i],
                         "No Exact Match|")
  yob = filter(wikiResults.items,
               yob==bgb3$birthDate[i]|yod==bgb3$deathDate[i])
  if (dim(yob)[1]>0) {
    bgb3$score[i] = paste0(bgb3$score[i],
                           "yob or yod Match|")
    bgb3[i,c("lopt","options","wdid")] = matchName(yob,bgb3[i,])
    if (bgb3$options[i]!="") {
      print(i)
      next
    }
  }
  #then try to match year of birth and death to wikidata items with floruit dates
  #15 years to account for childhood
  yob = wikiResults.floruit
  if (!is.na(bgb$birthDate[i])) {
    yob = filter(yob,
                 floruitDate1>bgb3$birthDate[i]+15)
  }
  if (!is.na(bgb$deathDate[i])) {
    yob = filter(yob,
                 floruitDate2<bgb3$deathDate[i])
  }
  if (dim(yob)[1]>0&dim(yob)[1]!=dim(wikiResults.floruit)[1]) {
    bgb3$score[i] = paste0(bgb3$score[i],
                           "floruit Match WD|")
    bgb3[i,c("lopt","options","wdid")] = matchName(yob,bgb3[i,])
    if (bgb3$options[i]!="") {
      print(i)
      next
    }
  }
  #finally, compare wikidata year of birth and death to inferred floruit dates from specimens
  #minSpecDate2 already includes a 15y correction for childhood
  if (is.na(bgb3$minSpecDate2[i])==F) {
    yob = filter(wikiResults.items,
                 is.na(yob)==F|is.na(yod)==F)
    yob = filter(wikiResults.items,
                 is.na(yob)|yob<bgb3$minSpecDate2[i])
    yob = filter(yob,is.na(yod)|yod>bgb3$maxSpecDate[i])
    if (dim(yob)[1]>0) {
      bgb3$score[i] = paste0(bgb3$score[i],
                             "floruit match WD-BGB|")
      bgb3[i,c("lopt","options","wdid")] = matchName(yob,bgb3[i,])
      print(i)
    }
  }
}

```

